**Red Wine Exploration**
=======================
***Jake Neyer\
January 14, 2018***
   

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(knitr)
library(GGally)
library(gridExtra)
library(nnet)
```

```{r echo=FALSE, Load_the_Data}
# Loading the Data
wine_data <- read.csv('/Users/jacobneyer/Desktop/EDA-with-R-master/wineQualityReds.csv')

```
## **Taking a Peek at the Dataset**
    
    

```{r echo=FALSE, message=FALSE, warning=FALSE}

str(wine_data)

```

## **Dataset Metadata**

    

  **Input variables (based on physicochemical tests):**
    
   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do \
   not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of\
   levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' and\
   flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, it's\
   rare to find wines with less than 1 gram/liter and wines with greater than 45 \
   grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between \
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial \
   growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentrations,\
   SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2\
   becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the \
   percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)\
   to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)\
   levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine


   **Output variable (based on sensory data):** 
   
   12 - quality (score between 0 and 10)

    
    

And a more human-readable version:

```{r kable, results='asis',echo=FALSE, message=FALSE, warning=FALSE}
kable(head(wine_data), format = 'markdown', width = 8)
```
```{r echo=FALSE}

wine_data$score[5 >= wine_data$quality] = "Bad"
wine_data$score[5 < wine_data$quality] = "Good"
wine_data$score[7 <= wine_data$quality] = "Great"
wine_data$score = as.factor(wine_data$score)

wine_data$score = relevel(wine_data$score, "Bad")

```



## **Univariate Plots Section**
    

Checking out the acidity of the wines:


```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity}
p1 <- qplot(data = wine_data, x = fixed.acidity, 
            geom = 'histogram', binwidth = .1)
p2 <- qplot(data = wine_data, x = log(volatile.acidity), 
            geom = 'histogram', binwidth = .05)
p3 <- qplot(data = wine_data, x = log10(citric.acid), 
            geom = 'histogram', binwidth = .05)
p4 <- qplot(data = wine_data, x = pH, geom = 'histogram', 
            binwidth = .01)

grid.arrange(p1,p2,p3,p4, ncol=2)

```

Barring citric acid content, the acidity seems to generally be normally distrobuted.\
Fixed acidity is centered around 6-8g/dm^3 and violatile acidity is generally \
between 0.4 and 0.7g/dm^3. I will look more into this in the following section. \
In order to normalize volatile acid and citric acid contents, I performed a \
logrithmic transformations on the two variables.
    
    
    
Taking a look at the other factors:


```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(data = wine_data, x = residual.sugar,
      geom = 'histogram', binwidth = .1)+
  scale_x_continuous(limits = c(0,10))

summary(wine_data$residual.sugar)

```
Residual sugar content is normally distrobuted around 2.2 g/dm^3.



```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(data = wine_data, x = chlorides, 
            geom = 'histogram', binwidth = .005)

summary(wine_data$chlorides)

```
    
  Chloride content is normally distributed around 0.0874 g/dm^3.  
    
    
```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(data = wine_data, x = log(free.sulfur.dioxide), 
            geom = 'histogram', binwidth = .1)

summary(wine_data$free.sulfur.dioxide)
```
     
In order to normalized the distrobution, I performed a logrithmic transformation\
on the free SO2 content. It appears to be for the most part normally distributed \
around 14/15 g/dm^3
     
```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(data = wine_data, x = log(total.sulfur.dioxide), 
            geom = 'histogram', binwidth = .05)

summary(wine_data$total.sulfur.dioxide)

```    

The total SO2 content is generally around  46 g/dm^3.
    
```{r echo=FALSE, message=FALSE, warning=FALSE}  
 
qplot(data = wine_data, x = sulphates, 
             geom = 'histogram', binwidth = .01)   

summary(wine_data$sulphates)

```
Sulphate content is normally distrobuted around .6 g/dm^3.


```{r echo=FALSE, message=FALSE, warning=FALSE}  

qplot(data = wine_data, x = density, 
            geom = 'histogram', binwidth = .0001)


summary(wine_data$density)
```
The density of the wines is normally distrobuted aroung .996 g/dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(data = wine_data, x = log(alcohol), 
             geom = 'histogram', binwidth = .01)

summary(wine_data$alcohol)
```

I used a logrithmic transformation to try to normalize the alcohol content, but \
it is still skewed significantly to the right.


    
Taking a look at quality distrobutions:

```{r echo=FALSE, message=FALSE, warning=FALSE}
p12 <- qplot(data = wine_data, x = quality, 
             geom = 'histogram', binwidth = 1)
p13 <- qplot(data = wine_data, x = "", y = quality, 
             geom = 'boxplot')

grid.arrange(p12,p13, ncol=2)
```
```{r echo=FALSE}

summary(wine_data$quality)

```

Quality ratings of 5 and 6 are extremely common; more so than any other. It is \
very rare for ratings to be 7 or above.



Taking a look at the modified quality distrobutions (Bad, Good, and Great):

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality}

qplot(data = wine_data, x = score)+
  geom_bar()

```

'Bad' rated wines are the most frequent, 'Good' wines are in close second, and \
there are even fewer 'Great' wines in the dataset.


## **Univariate Analysis**

```{r echo=FALSE}

summary(wine_data)

```


The dataset contains 1599 red wines with 13 features. Quality is rated on a scale \
from 0 (Very Bad) to 10 (Very Excellent). The most common quality ratings are 5 \
and 6 in this particular dataset. I wonder which features play the largest role \
in determing the quality rating. I am also curious to better understand how \
different variables affect others, such as alcohol content and density or acidity\
and residual sugar. In the following section I will begin to understand how the \
different variables interact and also understand which variables play roles in \
determining the overall quality of the wine.









## **Bivariate Plots Section**

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggpairs(wine_data[2:12], width = 14, height = 10,
         upper = list(continuous = wrap("cor", size = 3)))

```

Citric acid seems to be  correlated with many other variables in this dataset. \
Density also seems to be correlated with many other variables.



Taking a look at acidity as it relates to other variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_point()+
  geom_smooth(method = 'lm')


cor(wine_data$citric.acid, wine_data$fixed.acidity)

```

Citric acid has a strong, positive relationship with fixed acidity.



```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_point()+
  geom_smooth(method = 'lm')

cor(wine_data$citric.acid, wine_data$volatile.acidity)

```


Citric acid has a strong, negative relationnship with volatile acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_data, aes(x = density, y = fixed.acidity))+
         geom_jitter(alpha=1/3)+
  geom_smooth(method = 'lm')

cor(wine_data$density, wine_data$fixed.acidity)

```
Density has a strong, positive relationship with the fixed acidity.



```{r echo=FALSE, message=FALSE, warning=FALSE}

 ggplot(aes(x = alcohol, 
           y = density   ), 
       data = wine_data) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)   

cor(wine_data$alcohol, wine_data$density)

```

Density has a negative relationship with alcohol content. Generally speaking, \ 
wines with a higher alcohol content have a lower density.This makes sense because\
alcohol is less dense than water.



```{r echo=FALSE, message=FALSE, warning=FALSE}


ggplot(data = wine_data , aes(y = alcohol, x = factor(quality))) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

cor(wine_data$alcohol, wine_data$quality)

```

This shows not only the frequency of each quality rating, but also the statistical\
breakdown. The box plots give a good indication of what alcohol content each of the\
quality ratings have.




## **Bivariate Analysis**


I took a look at how the variables interacted with each other. Acidity clearly \
plays a large role in the wine. Both fixed and volatile acidity have strong \
relationships with many of the variables. The strongest relationship, maybe not \
so suprisingly is between citric acid and the fixed acidity of the wine. Another\
rather strong relationship is one between alcohol content and the quality of the\
wine.


Here is a detailed breakdown of the input variables relationship with the wine \
quality:

```{r echo=FALSE}

cor(wine_data$quality, subset(wine_data[2:12]))

```




## **Multivariate Plots**


```{r echo=FALSE, Multivariate_Plots}

p17 <- ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_jitter()+
    aes(color = quality)

p18 <- ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_jitter()+
    aes(color = quality)

grid.arrange(p17,p18)

```


The quality scale is difficult to visualize. Breaking the ratings into larger \
levels will help visualize the data more easily.




Running it again I get this:


```{r echo=FALSE}

p19 <- ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_jitter()+
  aes(color = score)


p20 <- ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_jitter()+
  aes(color = score)


grid.arrange(p19,p20)

```


As mentioned before, there is a stong relationship between acidity and citric acid\
content. Additionally, great wines seem to have less violatile acidity and higher\
fixed acidity.



```{r echo=FALSE}

ggplot(data = wine_data, aes(x = alcohol, y = sulphates, 
                            color = factor(quality))) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))


```


Better quality wines seem to have more sulphates. Additionally, lower quality wines\
are concentrated in the area of low sulphate content and low alcohol content.




```{r echo=FALSE}

ggplot(wine_data, aes(x = sulphates, y = pH))+
  geom_point(position = 'jitter', alpha = 1/2)+
  aes(color = score)+
  coord_cartesian(xlim = c(.3,1))+
  geom_smooth(method = "lm", se = FALSE,size=1)



```

PH levels do not seem to have much of an effect on wine quality, but once again\ 
we see sulphates do seem to play a role in the quality rating.




```{r echo=FALSE}

ggplot(wine_data, aes(x = alcohol, y = quality))+
  geom_count(position = 'jitter')+
  aes(color = score)


```

Lower quality wines tend to have lower alcohol content. This is an interesting\ 
visualization because it reiterates the size of each quality rating in a different\
way.



## **Multivariate Analysis**

Alcohol content clear plays one of the largest roles in determing the quality of\
the wines. Additional variables such as volatile acidity, fixed acidity, sulphate \
content and citric acid content also play a slightly less significant role in determining\
the quality.




### **Wine Quality Bar Graph**

```{r echo=FALSE, Plot_One}

 ggplot(data = wine_data, aes(x = as.factor(quality))) + 
  geom_bar(fill = 'darkred')+ 
  xlab("Wine Quality")

```

This plot displays the quality distrobution of the wine quality in the dataset \
and gives good insight on the percentage breakdown of the dataset based on quality.



### **Sulphates and Alcohol Content**
```{r echo=FALSE, Plot_Two}

ggplot(wine_data, aes(x = alcohol, y = sulphates, color = factor(quality)))+
  geom_point(position = 'jitter', alpha = 1/2)+
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
  coord_cartesian(ylim = c(.4,1.2))+
  ggtitle("Sulphates(g/dm^3) vs Alcohol Content(%), Stratified by Quality")+
  xlab("Alcohol (% by volume)")+
  ylab("Sulphate (g/dm^3)")



```


Wines with low sulphate content and low alcohol percentage are almost certainly \
low quality wines.



### **Alcohol Content and Quality**
```{r echo=FALSE, Plot_Three}


ggplot(wine_data, aes(x = alcohol, y = quality, color = factor(quality)))+
  geom_count(position = 'jitter', alpha = 1/2)+
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
  ggtitle("Alcohol Percentage and Quality")+
  xlab("Alcohol (% by volume)")+
  ylab("Quality of Wine")



```



This plot shows the clear stratification of wine quality based on alcohol content.\
I chose this as my final plot because it highlights the fact the alcohol content \
is a large determining factor in the wine's quality.




## **Reflection**

Through my exploratory data analysis performed on the 'Wine QUality' dataset, I \
discovered key variables that determine the quality of a wine. I identified alcohol\
as playing one of the largest roles in determining quality. Additionally, fixed and\
volatile acidity also play key roles in determining the quality. A larger dataset\
would be nice to explore the variables in more detail. Moving ahead, I would like\
to apply some different models to this dataset to run quality tests based on built-out\
algorithms.



## **Sources**

 P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib
                
                
https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/diamondsExample_2016-05.html

https://d17h27t6h515a5.cloudfront.net/topher/2017/February/58af99ac_projecttemplate/projecttemplate.rmd


----------------------
