**Red Wine Exploration**
=======================
***Jake Neyer\
January 14, 2018***
   

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(knitr)
library(GGally)
library(gridExtra)
library(nnet)
```

```{r echo=FALSE, Load_the_Data}
# Loading the Data
wine_data <- read.csv('/home/jake/Downloads/wineQualityReds.csv')

```
## **Taking a Peek at the Dataset**
    
    

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(wine_data)
```

## **Dataset Metadata**

    

  **Input variables (based on physicochemical tests):**
    
   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do \
   not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of\
   levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' and\
   flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, it's\
   rare to find wines with less than 1 gram/liter and wines with greater than 45 \
   grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between \
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial \
   growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentrations,\
   SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2\
   becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the \
   percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)\
   to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)\
   levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine


   **Output variable (based on sensory data):** 
   
   12 - quality (score between 0 and 10)

    
    

And a more human-readable version:

```{r kable, results='asis',echo=FALSE, message=FALSE, warning=FALSE}
kable(head(wine_data), format = 'markdown', width = 8)
```


## **Univariate Plots Section**
    

Checking out the acidity of the wines:


```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity}
p1 <- qplot(data = wine_data, x = fixed.acidity, geom = 'histogram', binwidth = .1)
p2 <- qplot(data = wine_data, x = log(volatile.acidity), geom = 'histogram', binwidth = .05)
p3 <- qplot(data = wine_data, x = log10(citric.acid), geom = 'histogram', binwidth = .05)
p4 <- qplot(data = wine_data, x = pH, geom = 'histogram', binwidth = .01)

grid.arrange(p1,p2,p3,p4, ncol=2)

```

Barring citric acid content, the acidity seems to generally be normally distrobuted.\
Fixed acidity is centered around 6-8g/dm^3 and violatile acidity is generally between \
0.4 and 0.7g/dm^3. I will look more into this in the following section. In order to normalize \
volatile acid and citric acid contents, I performed a logrithmic transformations on \
the two variables.
    
    
    
Taking a look at the other factors:


```{r echo=FALSE, message=FALSE, warning=FALSE, OtherFactors}

p5 <- qplot(data = wine_data, x = residual.sugar, geom = 'histogram', binwidth = .1)+
  scale_x_continuous(limits = c(0,10))
p6 <- qplot(data = wine_data, x = chlorides, geom = 'histogram', binwidth = .005)
p7 <- qplot(data = wine_data, x = log(free.sulfur.dioxide), geom = 'histogram', binwidth = .1)
p8 <- qplot(data = wine_data, x = log(total.sulfur.dioxide), geom = 'histogram', binwidth = .05)
p9 <- qplot(data = wine_data, x = density, geom = 'histogram', binwidth = .0001)
p10 <- qplot(data = wine_data, x = sulphates, geom = 'histogram', binwidth = .01)
p11 <- qplot(data = wine_data, x = log(alcohol), geom = 'histogram', binwidth = .01)

grid.arrange(p5,p6,p7,p8,p9,p10,p11, ncol = 3)


```


In order to get a more normally distributed graph of SO2 and alcohol content, I \
performed a few more logrithmic transformations.
    
     
    
Taking a look at quality distrobutions:

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality}
p12 <- qplot(data = wine_data, x = quality, geom = 'histogram', binwidth = 1)
p13 <- qplot(data = wine_data, x = "", y = quality, geom = 'boxplot')

grid.arrange(p12,p13, ncol=2)
```
```{r echo=FALSE}

summary(wine_data$quality)

```

Quality ratings of 5 and 6 are extremely common; more so than any other.


## **Univariate Analysis**

```{r echo=FALSE}

summary(wine_data)

```


The dataset contains 1599 red wines with 13 features. Quality is rated on a scale \
from 0 (Very Bad) to 10 (Very Excellent). The most common quality ratings are 5 and \
6 in this particular dataset. I wonder which features play the largest role in determing\
the quality rating. I am also curious to better understand how different variables \
affect others, such as alcohol content and density or acidity and residual sugar.\
In the following section I will begin to understand how the different variables interact\
and also understand which variables play roles in determining the overall quality of\
the wine.









## **Bivariate Plots Section**

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggpairs(wine_data[2:12], width = 14, height = 10,
         upper = list(continuous = wrap("cor", size = 3)))

```



```{r echo=FALSE, message=FALSE, warning=FALSE}

p14 <- ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_point()+
  geom_smooth(method = 'lm')


p15 <- ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_point()+
  geom_smooth(method = 'lm')

p16 <- ggplot(data = wine_data, aes(x = density, y = volatile.acidity))+
         geom_jitter(alpha=1/5)
  
  
grid.arrange(p14,p15)
grid.arrange(p16)


```



```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_data, aes(x = alcohol, y = density))+
         geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')


```





```{r echo=FALSE, message=FALSE, warning=FALSE}


ggplot(data = wine_data, aes(x = alcohol, y = quality))+
  geom_count(position = 'jitter', alpha = 1/2)

```



## **Bivariate Analysis**


I took a look at how the variables interacted with each other. Acidity clearly \
plays a large role in the wine. Both fixed and volatile acidity have strong relationships\
with many of the variables. The strongest relationship, maybe not so suprisingly \
is between citric acid and the fixed acidity of the wine. Another rather strong relationship\
is one between alcohol content and the quality of the wine.


Here is a detailed breakdown of the input variables relationship with the wine quality:

```{r echo=FALSE}

cor(wine_data, wine_data$quality)

```




## **Multivariate Plots**


```{r echo=FALSE, Multivariate_Plots}

p17 <- ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_jitter()+
    aes(color = quality)

p18 <- ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_jitter()+
    aes(color = quality)

grid.arrange(p17,p18)

```


The quality scale is difficult to visualize. Breaking the ratings into larger levels\
will help visualize the data more easily.

```{r echo=FALSE}

wine_data$score[5 >= wine_data$quality] = "Bad"
wine_data$score[5 < wine_data$quality] = "Good"
wine_data$score[7 <= wine_data$quality] = "Great"
wine_data$score = as.factor(wine_data$score)

wine_data$score = relevel(wine_data$score, "Bad")

```



Running it again I get this:


```{r echo=FALSE}

p19 <- ggplot(data = wine_data, aes(x = citric.acid, y = fixed.acidity))+
         geom_jitter()+
  aes(color = score)


p20 <- ggplot(data = wine_data, aes(x = citric.acid, y = volatile.acidity))+
         geom_jitter()+
  aes(color = score)


grid.arrange(p19,p20)

```


As mentioned before, there is a stong relationship between acidity and citric acid\
content. Additionally, great wines seem to have less violatile acidity and higher\
fixed acidity.



```{r echo=FALSE}

ggplot(wine_data, aes(x = alcohol, y = sulphates))+
  geom_point(position = 'jitter', alpha = 1/2)+
  aes(color = score)+
  coord_cartesian(ylim = c(.4,1.2))


```


Better quality wines seem to have more sulphates.




```{r echo=FALSE}

ggplot(wine_data, aes(x = sulphates, y = pH))+
  geom_point(position = 'jitter', alpha = 1/2)+
  aes(color = score)+
  coord_cartesian(xlim = c(.3,1))


```





```{r echo=FALSE}

ggplot(wine_data, aes(x = alcohol, y = quality))+
  geom_count(position = 'jitter')+
  aes(color = score)


```

Lower quality wines tend to have lower alcohol content.



## **Multivariate Analysis**

Alcohol content clear plays one of the largest roles in determing the quality of\
the wines. Additional variables such as volatile acidity, fixed acidity, sulphate \
content and citric acid content also play a slightly less significant role in determining\
the quality.




### **Wine Quality Bar Graph**

```{r echo=FALSE, Plot_One}

 ggplot(data = wine_data, aes(x = as.factor(quality))) + 
  geom_bar(fill = 'darkred')+ 
  xlab("Wine Quality")

```

This plot displays the quality distrobution of the wine quality in the dataset \
and gives good insight on the percentage breakdown of the dataset based on quality.



### **Sulphates and Alcohol Content**
```{r echo=FALSE, Plot_Two}

ggplot(wine_data, aes(x = alcohol, y = sulphates))+
  geom_point(position = 'jitter', alpha = 1/2)+
  aes(color = score)+
  coord_cartesian(ylim = c(.4,1.2))+
  ggtitle("Sulphates vs Alcohol Content, Stratified by Quality")+
  xlab("Alcohol (% by volume)")+
  ylab("Sulphate (g/dm^3)")



```


Wines with low sulphate content and low alcohol percentage are almost certainly \
low quality wines.



### **Alcohol Content and Quality**
```{r echo=FALSE, Plot_Three}


ggplot(wine_data, aes(x = alcohol, y = quality))+
  geom_count(position = 'jitter', alpha = 1/2)+
  aes(color = score)+
  ggtitle("Alcohol Percentage and Quality")+
  xlab("Alcohol (% by volume)")+
  ylab("Quality of Wine")



```



This plot shows the clear stratification of wine quality based on alcohol content.\
I chose this as my final plot because it highlights the fact the alcohol content \
is a large determining factor in the wine's quality.




## **Reflection**

Through my exploratory data analysis performed on the 'Wine QUality' dataset, I \
discovered key variables that determine the quality of a wine. I identified alcohol\
as playing one of the largest roles in determining quality. Additionally, fixed and\
volatile acidity also play key roles in determining the quality. A larger dataset\
would be nice to explore the variables in more detail. Moving ahead, I would like\
to apply some different models to this dataset to run quality tests based on built-out\
algorithms.



## **Sources**

 P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib
                
                
https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/diamondsExample_2016-05.html

https://d17h27t6h515a5.cloudfront.net/topher/2017/February/58af99ac_projecttemplate/projecttemplate.rmd


----------------------
